Detecting Stylistic Variation in Pop Production

### 2023-11-30 outputs/paper/paper.qmd
intro edits

We are interested in understanding how to create an approximation for production style. To do this we used Spotify API's audio features dataset to explore different audio feature metrics and identify which were the most import predictors in identifying pop producers' style. We focus on Jack Antonoff, as he is a prolific and widely used producer, who also has his own solo acts.

@@

We are interested in whether Spotify audio features can be used to train a model that distinguishes whether a song was produced by Antonoff. We obtained tracks from six artists who have collaborated with Antonoff, both before and during their collaboration. We also obtained tracks from similarly prolific pop producers: Joel Little, Ariel Rechstaid, Max Martin, Greg Kurstin, Paul Epworth and Rick Nowels. Our data set contains 3738 tracks in total, which were split into a training set (2802 tracks) and a testing set (936 tracks).

We built two models to explore the data. The first model, a logistic regression, draws on a dataset of artists who have collaborated with Antonoff (1954 tracks total), with a binary variable to tag whether the song was an Antonoff production. The training set included 1465 tracks, and the testing set included 489 tracks. With this model, we explore which variables are stronger predictors of Antonoff's style, developing a better understanding of whether the Spotify data can trace Antonoff's involvement in an artist's sound over time. We built a second model, a Random Forest (RF), to complement the insights from the logistic regression. The RF trains on a training set that includes other major pop producers in order to classify tracks based on who produced them, and return information on which predictors were most useful in doing so.

We find that `danceability` is the strongest predictor of whether a song has been produced by Antonoff. Furthermore, when `danceability`'s correlation scores with other features in the dataset are considered, the variance in correlation scores tends to be strongest classifier of whether a song was produced by Antonoff. In other words, danceability in concert with other features in the dataset provides the best predictions.

Our results suggest that we can capture something about Antonoff's music with these variables. It might, in the end, just be that subjective response that some people feel, and others enjoy, and that we are all just trying to describe in using different methods and tools. With the understanding that what we understand as "data-driven" in recommendation systems is often intertwined with human subjectivity -- engineers' understandings of taste and taste-making shape the recommendation systems that they build (@citeSeaver) -- we contribute to a more informed understanding of how style and tastes can been algorithmically rendered using Spotify's audio features.

>>>>

Style has been extensively studied in the field of computational stylistics, with researchers applying methods ranging from (x, y, z) to define and attribute aesthetics and patterns to artists.

We are interested in understanding how to create an approximation for production style using a data set built using the Spotify audio features API. To do this, we explore different audio feature metrics and identify which were the most import predictors in identifying pop producers' style. We focus on Jack Antonoff, as he is a prolific and widely used producer, who also has his own solo acts.

@@

We obtained tracks from six artists with extensive discographies who have collaborated with Antonoff, both before and during their collaboration. We also obtained tracks from similarly prolific pop producers: Joel Little, Ariel Rechstaid, Max Martin, Greg Kurstin, Paul Epworth and Rick Nowels. Our data set contains 3738 tracks in total, which were split into a training and test sets.

We built two models to explore the data. The first model, a logistic regression, is employed to observe any patterns in audio feature scores across Antonoff's collaborations with artists. The visualizations and analysis offer a closer look at Antonoff's oeuvre, along with a portrait of the trends in female singer-songwriter contemporary pop (according to Spotify's API). It draws on a dataset of artists who have collaborated with Antonoff (1954 tracks total), with a binary variable to tag whether the song was an Antonoff production. The training set included 1465 tracks, and the testing set included 489 tracks.

The second model, a Random Forest (RF), trains on a data set that includes the discographies of other prominent pop producers to classify producers. This expands upon and complements the results of the logistic regression -- where a logistic regression predicting an Antonoff-produced track in an artist's discography might actually be picking up on an artist releasing pop music, training a Random Forest on a bigger dataset with more producers captures both non-linear relationships in the data and accounts for intra-genre differences. In obtaining the strongest predictor variables in the Random Forest, we can better understand how it distinguishes between different pop producers' styles.

We find that, in both models, `danceability` is the strongest predictor of whether a song has been produced by Antonoff. Additionally, when we look at how danceability relates to other features in the dataset, we find that the differences in these relationships are very informative. Specifically, the degree to which danceability's relationships with other features vary helps us most accurately determine if Antonoff produced a song. Put simply, using danceability along with its interactions with other dataset features gives us the most accurate predictions.

In an interview with Spotify principal engineer Glen McDonald, we learned that the danceability and valence features were built by Spotify using crowdsourced methods.

Our results suggest that we can capture something about Antonoff's music with these variables. It might, in the end, just be that subjective response that some people feel, and others enjoy, and that we are all just trying to describe in using different methods and tools.

With the understanding that what we understand as "data-driven" in recommendation systems is often intertwined with human subjectivity -- engineers' understandings of taste and taste-making shape the recommendation systems that they build (@citeSeaver) -- we contribute to a more informed understanding of how style and tastes can been algorithmically rendered using Spotify's audio features.
